---
title: "Lab 8: ANOVA, chi-square, beeswarm, & cool table"
author: "Keene Morrow"
date: "11/19/2019"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(tidyverse)
library(car)
library(janitor)
library(ggbeeswarm)
library(kableExtra)
```

# 1. Read in the *penguins.csv* dataset
```{r}
penguins <- readr::read_csv("penguins.csv") %>% 
  janitor::clean_names()
```

### 2. Make a table of counts by species and sex

```{r}
penguin_counts <- penguins %>% 
  count(species, sex)

penguin_counts
```

# 3. Compare flipper sizes for female penguins across 3 species

Factor: species
Variable: flipper length


```{r}
# make subset of female penguins
penguin_f <- penguins %>%
  dplyr::filter(sex == "FEMALE")

# explore data with historgram
ggplot(data = penguin_f, aes(x = flipper_length_mm)) +
  geom_histogram() +
  facet_wrap(~species)

# explore data with QQ plot
ggplot(data = penguin_f, aes(sample = flipper_length_mm)) +
  geom_qq() +
  facet_wrap(~species)


```
# 4. Find summary statistics

```{r}

penguin_f_summary <- penguin_f %>%
  group_by(species) %>%
  summarize(
    mean_flipper_length = mean(flipper_length_mm),
    sd_flipper_length = sd(flipper_length_mm),
    sample_size = n(), # not count because count auto applies group_by
    se_flipper_length = sd(flipper_length_mm) / sqrt(n()),
    var_flipper_length = var(flipper_length_mm)
  )

penguin_f_summary

```

Largest group variance (`r round(max(penguin_f_summary$var_flipper_length), 2)`) is less than four times the smallest group variance (`r round(min(penguin_f_summary$var_flipper_length), 2)`), statisfies ANOVA assumption of equal variance.

Levene's Test (hypothesis test for equal variance)

```{r}
# h0: group variances are equal

leveneTest(flipper_length_mm ~ species, data = penguin_f)
```
Conclusion: Retain the null hyupothesis that variances are equal

All of these pieces combined suggest that one-wat ANOVA is appropriate was to compare means between the three species
- Normal distribution (hist, qq plot)
- Equal variances (variance calc, Levene's Test)

# 5. Make a beeswarm plot with mean and std deviation overlayed

Three most common error bars:
- standard deviation
- standard error
- confidence interval

```{r}
ggplot() +
  geom_beeswarm(data = penguin_f,
                aes(x = species,
                    y = flipper_length_mm),
                size = 1,
                alpha = 0.6,
                color = "grey60") +
  scale_x_discrete(labels = c("Adelie", "Chinstrap", "Gentoo")) +
  geom_point(data = penguin_f_summary,
             aes(x = species,
                 y = mean_flipper_length),
             color = "red",
             size = 2) +
  geom_errorbar(data = penguin_f_summary,
                aes(x = species,
                    ymin = mean_flipper_length - sd_flipper_length,
                    ymax = mean_flipper_length + sd_flipper_length),
                width = 0.1,
                color = "red") +
  theme_bw()

  
```
**NOTE** Make sure that if you're showing summary statistics, you actually say when you're showing and be consistent throughout

**Figure 1.** Flipper length (mm) for females of three penguin species (Adelie, Chinstrap, Gentoo) observed at Palm,er Station LTER, Antarctica.  Grey points indicate individual flipper lengths, red dots indicate group memes, error bars indicate $\pm$ 1 standard deviation. Data Source:

# 6. One-way ANOVA

Is there a significant difference in mean flipper length for female penguins between the three species (Adelie, Chinstrap, Gentoo)?

```{r}
penguin_aov <-  aov(flipper_length_mm ~ species, data = penguin_f)

summary(penguin_aov)

```
h0: all groups are drawn from populations with the same mean
hA: (at least two) means differ from each other

**High F value:** spread between groups is very large compared to the spread within each group
**Low p value (Pr(>F)):** It is very unlikely that these groups are drawn from populations with the same mean.

The mean flipper lengths for females of the three species are not equal.

Which ones differ signifcantly?

We need a post-hoc pairwaise comparison

# 7. Tukey's HSD
Tukey's is a pretty conservative post-hoc test

```{r}
TukeyHSD(penguin_aov)
```
We're most interested in the p adj (adjusted p value):

h0: no significant difference in means

Conclusion: One-way ANOVA with post-hoc Tueky's HSD reveals significant differences in mean flipper length across all species (Gentoo, mean = ; Chinstrap, mean = ; Adelie, mean = ) (F(df) = , p< 0.001, for all pairwise comparisons)

**NOTE** This alone is not terribly interesting wrt means comparison

# 7 . Making nicer tables (with proportions)

This is going to use vast simplications of the data for the purposes of learning
```{r}
party_relig <- gss_cat %>%
  filter(partyid %in% c("Strong republican", "Strong democrat")) %>%
  filter(relig %in% c("None", "Catholic")) %>%
  count(partyid, relig)

# Contingency table: uses two variables, one in rows, one in columns
party_relig_table <- party_relig %>%
  pivot_wider(names_from = relig, values_from = n)
```

# 8. Make a nice table of proportions using janitor::adorn_*

```{r}
party_relig_props <- party_relig_table %>%
  janitor::adorn_percentages(denominator = "row") %>% # direction of question matters!
  adorn_pct_formatting(digits = 1) %>%
  adorn_ns(position = "front")

party_relig_props # looks sad, bummer

kable(party_relig_props) %>%
  kable_styling()

```

- Are political and religious affiliations independent from each other?

-  Is there a significant association between political association and religious association?

- Is there a signficant effect of political association on religious affiliation? # careful here... this could imply causal direction)

- Is there a signifcant difference in religious affiliation between strong republicans and strong democrats?

Chi-squared
```{r}
# make contingency table tha tonly contains counts

chi_counts <- party_relig_table %>%
  select(-partyid)

chi_counts
```

h0:

- there is no significant association between party affiliation and relig aff
- there is not signifcant difference in relig aff between the two party groups
- party and relig aff are independent
(I'm a strong, independent party affiliation and I don't need no religious affiliation)

```{r}
my_party_chi <-  chisq.test(chi_counts)

my_party_chi

```

Conclusion:

- there is a significant difference in religious affiliation (none, Catholic) between strong republicans and strong dems ($\chi^2$ = ..., p < 0.001)

- there is significant association between party ID and religious affiliation

- party aff and relig aff are not independent
- 



